A family of accessibility measures

### 2024-11-12 paper/manuscript.qmd
finished Tij calculations using simple example

```{r load-packages, include=FALSE, cache=FALSE}
library(dplyr) # A Grammar of Data Manipulation
library(fitdistrplus) # Help to Fit of a Parametric Distribution to Non-Censored or Censored Data
library(geomtextpath) # Curved Text in 'ggplot2'
library(ggforce) # Accelerating 'ggplot2'
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(ggpmisc) # Miscellaneous Extensions to 'ggplot2'
library(ggrepel) # Automatically Position Non-Overlapping Text Labels with 'ggplot2'
library(ggspatial) # Spatial Data Framework for ggplot2
library(gridExtra) # Miscellaneous Functions for "Grid" Graphics 
library(here)

# library(huxtable) # Easily Create and Style Tables for LaTeX, HTML and Other Formats
# library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
# library(patchwork) # The Composer of Plots
# library(RColorBrewer) # ColorBrewer Palettes
# library(sf) # Simple Features for R
# library(scales) # Scale Functions for Visualization
# library(shadowtext) # Shadow Text Grob and Layer
# library(skimr) # Compact and Flexible Summaries of Data
# library(spdep) # Spatial Dependence: Weighting Schemes, Statistics
# library(tidyr) # Tidy Messy Data
# library(tmap) # Thematic Maps
```

@@

The following are simple numerical examples of each case assuming the following $T_{ij}$ matrix and a $f(c_{ij}) = \frac{1}{c_{ij}}$.

The variables:
```{r}
T_1 <- matrix(c(95, 23, 42, 27, 378, 45, 38, 36, 106), ncol = 3, byrow = TRUE)

T_1
```

```{r}
C <- matrix(c(2, 15, 5, 15, 2, 10, 5, 10, 2), ncol = 3, byrow = TRUE)

C
```

```{r}
O_i <- rowSums(T_1)
D_j <- colSums(T_1)

O_i
D_j
```

```{r}
T <- sum(sum(T_1))

T
```

## Unconstrained interaction case

$$
\hat{T_{ij}} = \frac{O_i\cdot D_j}{C_ij}
$$

```{r}
T_unc <- O_i %*% t(D_j) / C

T_unc
```

If we do the sums over rows, columns and all:
```{r}
rowSums(T_unc)
colSums(T_unc)
sum(sum(T_unc))
```

## Total constrained interaction case

Needs a single/global proportionality constant.
$$
\hat{T_{ij}} = k\frac{O_i\cdot D_j}{C_{ij}}
$$

```{r}
k <- T/sum(sum(T_unc))
k
```

```{r}
T_tot <- k * T_unc

T_tot
```

Let's check those marginals and total:
```{r}
rowSums(T_tot)
colSums(T_tot)
sum(sum(T_tot))
```

Destination constrained:
$$
\hat{T_{ij}} = \frac{B_j\cdot O_i\cdot D_j}{C_{ij}}
$$
```{r}
B_j <- colSums(T_1)/colSums(T_unc)

B_j <- matrix(rep(B_j, 3), ncol = 3, byrow = TRUE)

B_j
```

```{r}
T_dest <- B_j * T_unc

T_dest
```

Let's check those marginals and total:
```{r}
rowSums(T_dest)
colSums(T_dest)
sum(sum(T_dest))
```

Origin constrained:
$$
\hat{T_{ij}} = \frac{A_i\cdot O_i\cdot D_j}{C_{ij}}
$$

```{r}
A_i <- rowSums(T_1)/rowSums(T_unc)

A_i <- matrix(rep(A_i, 3), ncol = 3, byrow = FALSE)

A_i
```

```{r}
T_orig <- A_i * T_unc

T_orig
```

Let's check those marginals and total:
```{r}
rowSums(T_orig)
colSums(T_orig)
sum(sum(T_orig))
```

Unconstrained accessibility (no proportionality constant):
$$
S_i = \sum_j\frac{D_j}{C_{ij}}
$$

Total flow constrained accessibility, needs a balancing factor:
```{r}
k <- sum(D_j)/sum(S_i)

Total flow constrained accessibility:
Unconstrained accessibility (no proportionality constant):
$$
S_i = k\sum_j\frac{D_j}{C_{ij}}
$$

@@

Imagine if the interaction variable was in fact _potential_ interaction, and the units were trips but opportunities. We don't have to change much!

Referring to the preceding section, the entropy maximizing method can be used to derive an equivalent family of measures, but for the case of accessibility where the _interaction_ variable is $V_{ij}$, represent _potential interaction_, i.e., the number of potential reachable opportunities. $T_{ij}$ in @eq-phys-gravity-model can be replaced with $V_{ij}$ and presented as follows:

@@

Using the maximizing entropy approach... <!--DESCRIBING WHAT EACH MICRO/MESO/MACRO STATE IS HERE IN RELATION TO ACCESSIBILITY--> For each case...

## Unconstrained case

>>>>

```{r load-packages, include=FALSE, cache=FALSE}
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(flextable)
# library(huxtable) # Easily Create and Style Tables for LaTeX, HTML and Other Formats
# library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
# library(patchwork) # The Composer of Plots
# library(RColorBrewer) # ColorBrewer Palettes
# library(sf) # Simple Features for R
# library(scales) # Scale Functions for Visualization
# library(shadowtext) # Shadow Text Grob and Layer
# library(skimr) # Compact and Flexible Summaries of Data
# library(spdep) # Spatial Dependence: Weighting Schemes, Statistics
# library(tidyr) # Tidy Messy Data
# library(tmap) # Thematic Maps
# library(ggpmisc) # Miscellaneous Extensions to 'ggplot2'
# library(ggrepel) # Automatically Position Non-Overlapping Text Labels with 'ggplot2'
# library(ggspatial) # Spatial Data Framework for ggplot2
# library(gridExtra) # Miscellaneous Functions for "Grid" Graphics 
# library(here)
```

@@

The following are simple numerical examples of each case assuming the following 1) $T_{ij}$ trip ends matrix, 2) $C_{ij}$ travel cost matrix, and 3) an assumed travel impedance function of $f(c_{ij}) = \frac{1}{c_{ij}}$. For all cases, we look to estimate each $T_{ij}$ value using these three inputs, we present the known trip matrix and the trip ends below:
```{r}
known_trip_matrix <- matrix(c(95,23,42, 27,378,45, 38,36,106), ncol = 3, byrow = TRUE)
#unknowns <- matrix(c(rep(NA,9)), ncol=3, byrow=TRUE)
W_i <- known_trip_matrix |> rowSums()
W_j <- known_trip_matrix |> colSums()
T <- ifelse(sum(W_i)==sum(W_j), sum(W_j),NA)
knowns_longer <- cbind(known_trip_matrix,W_i)
knowns_longer <- rbind(knowns_longer,cbind(t(W_j),T))

rownames(knowns_longer) <- c("1","2","3", "TOTAL") 
colnames(knowns_longer) <- c("1","2","3", "TOTAL")

knowns_longer #|> as.data.frame() |> round() |> gt(rownames_to_stub=TRUE)
```

And the matrix of cost of travel from $i$ (rows) to $j$ (columns) is as follows:
```{r}
C <- matrix(c(2, 15, 5, 15, 2, 10, 5, 10, 2), ncol = 3, byrow = TRUE)
C_longer <-C
rownames(C_longer) <- c("1","2","3") 
colnames(C_longer) <- c("1","2","3") 
C_longer #|> as.data.frame() |> gt(rownames_to_stub=TRUE)
```

### Unconstrained interaction case

$k$ can be understood as being set to 1, so the model looks as follows, with no constraints and the trip matrix is as follows. For example, $T_{1,1} = 160*160*\frac{1}{2}$

$$
T_{ij} = W_i^{(1)} W_j^{(2)} f(c_{ij})
$$
```{r}
T_unc <- W_i %*% t(W_j) / C

W_i_unc <- T_unc |> rowSums()
W_j_unc <- T_unc |> colSums()
T_TOTAL_unc <- ifelse(sum(W_i_unc|>round())==sum(W_j_unc|>round()), sum(T_unc), NA)
T_unc_longer <- cbind(T_unc,W_i_unc)
T_unc_longer <- rbind(T_unc_longer, cbind(t(W_j_unc),T_TOTAL_unc))

rownames(T_unc_longer) <- c("1","2","3", "TOTAL") 
colnames(T_unc_longer) <- c("1","2","3", "TOTAL")

T_unc_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
```

As we can surmise, the trip ends do not add up to anything we know -- they do not add up to the known trip ends. Instead, they take on the assumptions of all balancing factors $k$ being equal to 1. When would this happen? It's hard to say... as the balancing factors themselves do not have practical meaning; what has meaning are the constraints. By satisfying the constraints, the model gains meaning. In this way, the resulting unconstrained interaction matrix is simply an expression of the magnitude of interaction of $W_i$ to $W_j$ assuming $f(c_{ij})$. In the context of spatial interaction modelling, this model is not entirely useful; it is in units of 'trips' but trips that satisify the macro constraint of $k$ equaling 1, which is practically meaningless.

### Total constrained interaction case

In the total constrained case, $k$ is replaced with $K$ (@eq-bf-K-gravitymodel satisfying only constraint @eq-constraint0-gravitymodel) so the model looks as follows:
$$
T_{ij}' = K W_i^{(1)} W_j^{(2)} f(c_{ij})
$$
\where Solved for our example, $K$ is equal to the total number of known trips divided by the sum of all the unconstrained trips made:
$$
K = \frac{T}{\sum_{i}\sum_{j} W_i^{(1)} W_j^{(2)} f(c_{ij})}\\
K = \frac{T}{\frac{W_1^{(1)}W_1^{(2)}}{c_{1,1}}+\frac{W_2^{(1)}W_1^{(2)}}{c_{2,1}}+\frac{W_3^{(1)}W_1^{(2)}}{c_{3,1}}+\frac{W_1^{(1)}W_2^{(2)}}{c_{1,2}} +... \frac{W_3^{(1)}W_3^{(2)}}{c_{3,3}}
}\\
K = \frac{790}{166443} \\
K = 0.00474636
$$

```{r}
K <- T/sum(sum(T_unc))
T_tot <- K * T_unc

W_i_tot <- T_tot |> rowSums()
W_j_tot <- T_tot |> colSums()
T_TOTAL_tot <- ifelse(sum(W_i_tot|>round())==sum(W_j_tot|>round()), sum(T_tot), NA)
T_tot_longer <- cbind(T_tot,W_i_tot)
T_tot_longer <- rbind(T_tot_longer, cbind(t(W_j_tot),T_TOTAL_tot))

rownames(T_tot_longer) <- c("1","2","3", "TOTAL") 
colnames(T_tot_longer) <- c("1","2","3", "TOTAL")

T_tot_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
```
While we can see that the total trips `r T` results, the total trip ends are not equal to the known trip ends as these constraints are not specified and thus not satisfied by the model.

### Singly-constrained

Two versions of this case, either origin-constrained or destination-constrained.

For the origin-constrained model, where $k$ is replaced with $A_i$ (@eq-bf-A-gravitymodel satisfying only constraint @eq-constraint1-gravitymodel and implicitly @eq-constraint0-gravitymodel), the model looks as follows:
$$
T_{ij}^{''(1)} = A_i W_i^{(1)} W_j^{(2)} f(c_{ij})
$$
\where In our example, $A_i$ is:
$$
A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}\\
A_1 = \frac{1}{W_1^{(2)} f(c_{1,1}) + W_2^{(2)} f(c_{1,2}) + W_3^{(2)} f(c_{1,3})}\\
...\\
A_1 = \frac{1}{160/2+437/15+193/5} =0.006768953 \\
A_2 = \frac{1}{160/15+437/2+193/10} =0.004024685 \\
A_3 = \frac{1}{160/5+437/10+193/2} =0.005807201 
$$

```{r}
A_i <- 1/t((matrix(W_j))) %*% (1/C)
#A_i <- rowSums(known_trip_matrix)/rowSums(T_unc) #our answer with knowns
A_i<- diag(A_i|> as.numeric(), nrow=3, ncol=3)

T_orig <-  A_i%*%T_unc

W_i_orig <- T_orig |> rowSums()
W_j_orig <- T_orig |> colSums()
T_TOTAL_orig <- ifelse(sum(W_i_orig)==sum(W_j_orig), sum(T_orig), NA)
T_orig_longer <- cbind(T_orig,W_i_orig)
T_orig_longer <- rbind(T_orig_longer, cbind(t(W_j_orig),T_TOTAL_orig))

rownames(T_orig_longer) <- c("1","2","3", "TOTAL") 
colnames(T_orig_longer) <- c("1","2","3", "TOTAL")

T_orig_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
```

We can see that the $W_i^{(1)}$ values are equal to the trip ends, but not $W_j^{(2)}$ values.

Similarly, the singly-constrained case can be flipped to be origin-constrained (@eq-bf-B-gravitymodel satisfying only constraint @eq-constraint1-gravitymodel and implicitly @eq-constraint0-gravitymodel):
$$
T_{ij}^{''(2)} = B_j W_i^{(1)} W_j^{(2)} f(c_{ij})
$$
\where $B_j$ is, and equal to:
$$
B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}\\
$$
```{r}
B_j <- 1/(t(matrix(W_i))) %*% (1/C)

B_j <- matrix(B_j) |> t()
B_j_longer <- B_j
colnames(B_j_longer) <- c("1","2","3")
B_j_longer  #|> as.data.frame()  |> gt()
#B_j <- colSums(known_trip_matrix)/colSums(T_unc) #our answer with knowns
```

And trips are:
```{r}
B_j <- diag(B_j|> as.numeric(), nrow=3, ncol=3)

T_dest <-  (B_j)%*%t(T_unc) |> t()

W_i_dest <- T_dest |> rowSums()
W_j_dest <- T_dest |> colSums()
T_TOTAL_dest <- ifelse(sum(W_i_dest)==sum(W_j_dest), sum(T_dest), NA)
T_dest_longer <- cbind(T_dest,W_i_dest)
T_dest_longer <- rbind(T_dest_longer, cbind(t(W_j_dest),T_TOTAL_dest))

rownames(T_dest_longer) <- c("1","2","3", "TOTAL") 
colnames(T_dest_longer) <- c("1","2","3", "TOTAL")

T_dest_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
```

And as we can observe, the $W_j^{(2)}$ values are equal to the known trip ends, while $W_i^{(1)}$ is not.

### Doubly-constrained case

The parameter $k$ is replaced with both $A_i$ and $B_j$ (@eq-bf-A-gravitymodel and @eq-bf-B-gravitymodel satisfying all three constraints @eq-constraint1-gravitymodel,  @eq-constraint2-gravitymodel and implicitly @eq-constraint0-gravitymodel), the model looks as follows:
$$
T_{ij}^{'''} = A_i B_j W_i^{(1)} W_j^{(2)} f(c_{ij})
$$
\where In our example, $A_i$ and $B_j$ is:
$$
A_i = \frac{1}{\sum_j B_j W_j^{(2)} f(c_{ij})}\\
B_j = \frac{1}{\sum_i A_i W_i^{(1)} f(c_{ij})}
$$
```{r}
doubly_constrained_gravity_model <- function(
    #CREDIT: I converted the python code from (https://github.com/SadraDaneshvar/Gravity_Model) to R
  O,  # Origin production weights vector
  D,  # Destination attraction weights vector
  impedance_matrix,  # the resulting travel cost impedance matrix
  error_threshold = 0.001,  # Error threshold for stopping condition
  improvement_threshold = 1e-6  # Improvement threshold for stopping condition
) {

  # Normalize O and D
  sum_O <- sum(O)
  sum_D <- sum(D)
  cat("Checking production, attraction balancing:\n")
  cat("Production: ", sum_O, "\n")
  cat("Attraction: ", sum_D, "\n")

  if (sum_O != sum_D) {
    cat("Productions and attractions do not balance, scaling to match.\n")
    if (sum_O < sum_D) {
      O <- O * (sum_D / sum_O)
    } else {
      D <- D * (sum_O / sum_D)
    }
  } else {
    cat("Production, attraction balancing OK.\n")
  }

  n <- length(O)  # Number of i
  T <- sum(O)  # Total trips
  Ai <- rep(1, n)  # Initialize Ai
  Bj <- rep(1, n)  # Initialize Bj

  previous_error <- Inf
  iteration_count <- 0
  stop_reason <- ""

  while (TRUE) {
    iteration_count <- iteration_count + 1

    # Update Ai
    for (i in 1:n) {
      Ai[i] <- 1 / (sum(Bj * D * impedance_matrix[i, ]) + 1e-9)
    }

    # Update Bj
    Bj_new <- rep(1, n)
    for (j in 1:n) {
      Bj_new[j] <- 1 / (sum(Ai * O * impedance_matrix[, j]) + 1e-9)
    }

    # Calculate Tij
    Tij <- outer(Ai * O, Bj_new * D) * impedance_matrix

    # Print Ai and Bj for this iteration
    cat("\nIteration:", iteration_count, "\n")
    cat("Ai:", round(Ai, 3), "\n")
    cat("Bj:", round(Bj_new, 3), "\n")

    # Calculate error
    error <- (sum(abs(O - rowSums(Tij))) + sum(abs(D - colSums(Tij)))) / T

    # Calculate change in error
    error_change <- abs(previous_error - error)

    # Check stopping conditions
    if (error < error_threshold) {
      stop_reason <- "Error threshold met"
      break
    } else if (error_change < improvement_threshold) {
      stop_reason <- "Slow improvement"
      break
    }

    previous_error <- error
    Bj <- Bj_new
  }

  # report stopping condition and final error threshold
  cat("\nStopping Condition: ", stop_reason, "\n")
  cat(sprintf("Final Error: %.3f%%\n", error * 100))

  return(list(Ai = Ai, Bj = Bj, Tij = Tij))
}

dc_results <- doubly_constrained_gravity_model(O=W_i, D=W_j, impedance_matrix = 1/C)
```

The $A_i$ and $B_j$ matrices:
```{r}
A_i_doubly <- dc_results$Ai
B_j_doubly <- dc_results$Bj

A_i_doubly |> matrix()
B_j_doubly |> matrix() |> t()
```

The trip matrix (notice, all trip ends are equivalent to the known trip ends)
```{r}
T_doubly <- dc_results$Tij

W_i_doubly <- T_doubly |> rowSums()
W_j_doubly <- T_doubly |> colSums()
T_TOTAL_dest <- ifelse(sum(W_i_doubly)==sum(W_j_doubly), sum(T_doubly), NA)
T_doubly_longer <- cbind(T_doubly,W_i_dest)
T_doubly_longer <- rbind(T_doubly_longer, cbind(t(W_j_doubly),T_TOTAL_dest))

rownames(T_doubly_longer) <- c("1","2","3", "TOTAL") 
colnames(T_doubly_longer) <- c("1","2","3", "TOTAL")

T_doubly_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
```

### Mean absolute percentage error (MAPE) between all cases

```{r}
#unconstrained
((sum(abs(W_i_unc - rowSums(known_trip_matrix))) + sum(abs(W_j_unc - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01)

#total constrained
((sum(abs(W_i_tot - rowSums(known_trip_matrix))) + sum(abs(W_j_tot - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01)

#singly-constrained (orig)
((sum(abs(W_i_orig - rowSums(known_trip_matrix))) + sum(abs(W_j_orig - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01)

#singly-constrained (dest)
((sum(abs(W_i_dest - rowSums(known_trip_matrix))) + sum(abs(W_j_dest - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01)

#doubly-constrained (dest)
((sum(abs(W_i_doubly - rowSums(known_trip_matrix))) + sum(abs(W_j_doubly - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01)
```

@@

Imagine if the interaction variable was in fact _potential_ interaction, and the units were trips but opportunities. We don't have to change much! Referring to the preceding section, the entropy maximizing method can be used to derive an equivalent family of measures, but for the case of accessibility where the _interaction_ variable is $V_{ij}$, represent _potential interaction_, i.e., the number of potential reachable opportunities. $T_{ij}$ in @eq-phys-gravity-model can be replaced with $V_{ij}$ and presented as follows:

@@

## Unconstrained accessibility 
(no proportionality constant):
$$
S_i = \sum_j\frac{D_j}{C_{ij}}
$$

##Total flow constrained accessibility, needs a balancing factor:
```{r}
k <- sum(D_j)/sum(S_i)

## Unconstrained accessibility (no proportionality constant):
$$
S_i = k\sum_j\frac{D_j}{C_{ij}}
$$
